---
title: "Report_Exercice1_Maelle"
output: html_document
date: "2025-03-05"
---
## Introduction 

**The data**

The data we have are used to assess the soil's response to an increase in the amount of CO2 in the air. They come from several experiments, for which the values were measured at different times. The different values are the carbon concentration in the air in g/m3, and the organic carbon concentration in the soil resulting from this.

**The importation**

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(lubridate)
library(tidyr)
library(readr)
library(stringr)
library(purrr)
library(knitr)

data <- readr::read_delim("../Data/Clean_data.csv",delim = ";",show_col_types = FALSE)
```

## The hypothesis

The logical assumption is that the concentration of carbon in the soil increases if the concentration in the air is higher, which means that the log ratio should be positive. 
But in addition to that time should have an influence. At first plants should grow more, and add a significant amount of carbon to the soil. The LRR should increase. But then other nutrients could become limiting and soil decomposition could accelerate. The LRR could stagnate or decrease.

## The calculation 

**Separation of the data in phases**

The column year is used to differenciate the phases

```{r}
early_phase <- data %>%
  select(year,ambient_CO2_mean, increased_CO2_mean) %>%
  filter(year<3)
mid_phase <- data %>%
  select(year,ambient_CO2_mean, increased_CO2_mean) %>%
  filter(year>=3 & year < 6)
late_phase <- data %>%
  select(year,ambient_CO2_mean, increased_CO2_mean) %>%
  filter(year>=6)
```

**Compute the log response ratio for every phases**

The log response ratio is calculated for soil organic carbon

```{r}
RR_early <- log(early_phase$increased_CO2_mean/early_phase$ambient_CO2_mean)
RR_mid <- log(mid_phase$increased_CO2_mean/mid_phase$ambient_CO2_mean)
RR_late <- log(late_phase$increased_CO2_mean/late_phase$ambient_CO2_mean)

Aggregated_RR_early <- mean(RR_early)
Aggregated_RR_mid <- mean(RR_mid)
Aggregated_RR_late <- mean(RR_late)
```

## The result
**The result :**
```{r echo=FALSE}
df <- data.frame(
  Name = c("RR early phase", "RR mid phase", "RR late phase"),
  Score = c(Aggregated_RR_early, Aggregated_RR_mid, Aggregated_RR_late)
)
kable(df, format = "markdown", caption ="log-response ratio")
```
**The interpretation**

The result are surprising, mostly because the RR of the early and late phase are negative. This could be explained because in the early phase, the plants did not have the time to adapt and response, then in the mid phase the plants grow more because of elevated CO2 (positive RR). And finally in the late phase decomposition is high and maybe limiting nutrients are limiting the growth, leading to negative RR again. 
